---
title: Activity Monitoring Project(1)
       Reproducible Research
author: "LonePine"
date: "April 5, 2016"
output: html_document
---
## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a **Fitbit, Nike Fuelband, or Jawbone Up**. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at *5 minute* intervals through out the day. The data consists of *two* months of data from an anonymous individual collected during the months of *October* and *November*, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

[Dataset:] (https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
Activity monitoring data [52K]
The **variables** included in this dataset are:

**steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)

**date**: The date on which the measurement was taken in YYYY-MM-DD format

**interval**: Identifier for the 5-minute interval in which measurement was taken.

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

**loading the required packages for the project**

*Read in the data*
```{r,echo = TRUE}
library(ggplot2)
library(plyr)
Activity <- read.csv("activity.csv")
head(Activity)
summary(Activity)
```

### processing the data
**convert date from factor to date, removing NA's and cleaning the data**
```{r, echo=TRUE}
Activity$date <- as.Date(Activity$date)

Activity_noNA <- na.omit(Activity)
head(Activity_noNA)
```

**finding the sum of steps taken each day and storing in a dataframe**

```{r, echo=TRUE}
dailysum <- aggregate(Activity_noNA$steps ~ Activity_noNA$date, FUN = sum)
colnames(dailysum) <- c("date","steps")
head(dailysum)
```

**Plotting a histogram of the total number of steps taken each day versus the date**

```{r, echo=TRUE}
hist(dailysum$steps, breaks = 20, xlab = "Steps", main = "Total Steps taken per day", col = "blue")
# mean of number of steps taken per day
mean(dailysum$steps)

#median of number of steps taken per day
median(dailysum$steps)
```

**Time series plot of the average number of steps taken**

```{r, echo=TRUE}
# calculate the average number of steps taken in each time interval in a day
averageSteps <- aggregate(Activity_noNA$steps, list(interval = as.numeric(Activity_noNA$interval)), FUN = "mean")
colnames(averageSteps) <- c("interval", "meanSteps")
head(averageSteps)
```

**Make a time series line plot for the average number of steps(X axis) Vs time interval(Y-axis)**

```{r,echo=TRUE}
ggplot(averageSteps, aes(interval, meanSteps)) + geom_line(color = "blue", size = 1.0) + 
       labs(x = "Time in 5 min intervals(min)", y = " Average number of steps for each 5 min.interval",             title = "Time Series Plot of average steps per interval") 
```

** The 5 minute interval that on average contains the maximum number of steps**

```{r, echo=TRUE}
MaximumSteps <- max(averageSteps$meanSteps)
averageSteps[averageSteps$meanSteps == MaximumSteps, 1]

```

**6. Code to describe and show a strategy for imputing missing data**

*First we find the number of rows with NA.In order to fill in values for missing data we could take the mean for that 5 minute interval and substitute*
```{r,echo=TRUE}
library(dplyr)
nrow(Activity[is.na(Activity$steps),])
# There are 2304 rows with steps = NA
Activity_noNA$day <- weekdays(as.Date(Activity_noNA$date))
# calculate the average number of steps per 5 minute interval and the day of the week
averageTable <- ddply(Activity_noNA, .(interval,day), summarize, 
                      meanSteps = mean(Activity_noNA$steps))
head(averageTable)
```

**Create data set with NA's for substituting with data with meanSteps**

```{r,echo= TRUE}
NAdata <- Activity[is.na(Activity$steps), ]
# substitute NA's in this dataset with the dataset having average number of steps per weekday per interval
newData <- Activity
for(i in 1:nrow(newData)){
 if (is.na(newData$steps[i])){
newData$steps[i] <- averageSteps[which(newData$interval[i] == averageSteps$interval), ]$meanSteps
 }
} 
head(newData)
```

**7. Histogram of the total number of steps taken each day after missing values are imputed**
```{r, echo=TRUE}
hist(newData$steps, breaks = 20, xlab = "steps taken per day", 
     main = "Total Steps taken each day after imputing missing values", col= "Yellow")
ggplot(newData, aes(date,steps)) + geom_bar(stat = "identity", colour = "yellow", width = 0.8)
# + labs( x = "Steps taken per day",
#  title = "Total steps taken each day after imputing missing values", y = "Total number of steps")
# Calculating the mean and median after imputing missing values with average for the day for a time interval
newDailySum <- aggregate(newData$steps ~ newData$date, FUN = sum)
colnames(newDailySum) <- c("date","steps")
mean(newDailySum$steps)
median(newDailySum$steps)
# Finding the difference between New and old mean and median 
meanDifference <- mean(newDailySum$steps) - mean(dailysum$steps)
medianDifference <- median(newDailySum$steps) - median(newDailySum$steps)
# There is no difference between the mean of new imputed data and old data
# There is no difference in the median between the old median and the new median
```
** 8. Panel plot comparing the average number of steps taken per 5-minute interval**
**    across weekdays and weekends**
** create a factor including weekend and weekdays in the dataset**
```{r, echo= TRUE}
newData$weekdays <- factor(format(newData$date, "%A"))
levels(newData$weekdays)
levels(newData$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday", 
                                             "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
# calculate the average steps across weekdays and weekends
weekdayAverage <- aggregate(newData$steps, list(interval = as.numeric(newData$interval), weekdays = newData$weekdays), FUN = "mean")
names(weekdayAverage)[3] <- "meanSteps"
library(lattice)
xyplot(weekdayAverage$meanSteps ~ weekdayAverage$interval | weekdayAverage$weekdays,
       layout = c(1,2), type = "l", xlab = "interval", ylab = "Number of steps")
```

```
  
```